\section{Contributions}

In this project, we develop and present a novel
MPST-based framework for developing full-stack
interactive TypeScript applications over WebSocket transport.

We motivate our API generation approach from 
\cite{Hybrid2016,PureScript2019} to generate TypeScript APIs
from a Scribble protocol specification.
By writing their full-stack applications using
the generated APIs, developers enjoy 
communication safety guarantees in their endpoint
applications by construction.
APIs generated for server-side endpoints
are compatible with the Node.js runtime,
as detailed in \cref{chap:node}.
For browser-side endpoints,
we present a novel approach for integrating
session types into web-based GUI programming based on
translating the theory on model types \cite{MVU2020}
to idiomatic practices on the React.js framework;
this is highlighted in \cref{chap:react}.
With respect to session type theory, implementations using
the generated APIs statically enjoy linear channel usage
guarantees and affine channel usage guarantees for back-end
and front-end targets respectively.

Compared to previous work 
\cite{Exceptional,PureScript2019,MVU2020,LINKS} on
session-typed web development, 
we are not only able to
statically provide the same level of communication safety
guarantees, but we do so
using modern web programming practices to increase
the relevance and usability of our work in industry:
our work targets TypeScript, Node.js and React.js,
and we explain in \cref{chap:ext} how we support advanced
web development idioms (such as asynchronous implementations)
and handle premature disconnections gracefully.

We also do not limit support to protocols that describe
a server-centric topology: we formalise \newtheory,
a theory of routed multiparty session types, in 
\cref{chap:theory} to prove that it is possible to relax the
server-centric topology assumption over WebSocket transport in a way 
that preserves communication along with the communication safety 
properties inherited from canonical session type theory.

% PLACES
Our work has received positive feedback from academia.
The initial part of this project was published as 
\emph{Generating Interactive WebSocket 
Applications in TypeScript} in Volume 314 of the Electronic Proceedings
in Theoretical Computer Science (EPTCS) \cite{PLACES2020},
in which I am the first and lead author.
The materials in \cref{chap:codegen,chap:node,chap:react}
are expanded from \cite{PLACES2020}.
The materials in 
\cref{chap:ext,chap:p2p,chap:theory,chap:impl,chap:eval}
are developed exclusively in this project by myself.
We were also invited to present a part of this work
at the 12th International Workshop on
Programming Language Approaches to Concurrency- \& 
Communication-cEntric Software (PLACES 2020).

% OVERALL
Overall, we offer an end-to-end solution for integrating
multiparty session type theory into modern full-stack web programming
practices, in a way that actively encourages developers to
design their application with communication safety in mind.
Our formalism of routed multiparty session types enables our
API generation tool to support a wider range of protocols, and
further reveals the potential
for implementing dynamic communication structures 
over centralised
network topologies.