\section{EFSM Encoding}
\label{section:reactefsm}

We encode each EFSM state as a React component.
This encoding is consistent with the semantics of model types,
as outlined in \cref{table:efsmmodeltypes}.
\fancyname{ReactMPST} generates an abstract React component class
for each EFSM state. The developer implements the generated API
by extending the abstract class to define their own
view function and implement any abstract methods
required by the EFSM state.

\renewcommand{\arraystretch}{1.5}
\begin{table}[!ht]
\centering
\begin{tabularx}{\linewidth}{||c|X||}
\hline
Model Type Property & \multicolumn{1}{c||}{Equivalent Abstraction for EFSM} \\
\hline\hline
View Function & 
Each EFSM state is a React component 
with its own \texttt{render()} method. \\ 
\hline
Set of Messages & The permitted channel actions are defined either as 
component props (for send states) 
or abstract instance methods (for receive states). \\
\hline
Update Function & EFSM state components are rendered by a session runtime,
so they notify the runtime to trigger a transition. \\
\hline
\end{tabularx}
\captionof{table}{Implementing Model Types as React Components}
\label{table:efsmmodeltypes}
\end{table}
\renewcommand{\arraystretch}{1}

Our EFSM encoding for browser-side endpoints remain
consistent with those for server-side endpoints, in the sense that
channel resources are abstracted away from the APIs generated
for the developer, so it is impossible to trigger
channel actions \textit{more than once} at any given EFSM state,
by construction.

\subsection{Send States}

For React components implementing send states,
we need to guarantee that the channel actions on the DOM
precisely correspond to the permitted transitions.
We require that send actions can only be invoked
by user-triggered events on UI elements, such as clicking a button or
typing in a textbox -- this generally aligns
with idiomatic front-end web development practices.

Without \fancyname{ReactMPST},
an implementation of state S40 from \cref{fig:adderclientfsm}
may resemble the following, where the developer attaches
event listeners on UI elements to trigger channel actions:

\begin{lstlisting}[language=javascript,numbers=none]
<button onClick={ev => (*@\hl{Add}@*)(this.state.num1, this.state.num2)}>
	Add
</button>
<button onMouseOver={ev => (*@\hl{Quit}@*)(this.state.message)}>
	Quit
</button>
\end{lstlisting}

If we directly provide the highlighted channel actions 
(i.e. via props), we cannot guarantee linear usage in the code 
without a linear type system.

What we want to work towards is something similar to the following,
where the highlighted component encapsulates the required information: 
\textbf{(1)} what event to react to, 
and \textbf{(2)} what payload to send when the event is triggered.
\texttt{<AddOnClick>} should guarantee, by construction,
that clicking on any child element will send the \tmsg{ADD}
message with the specified payload.

\begin{lstlisting}[language=javascript,numbers=none]
(*@\hl{<AddOnClick payload=\{[this.state.num1, this.state.num2]\}>}@*)
	<button>Add</button>
(*@\hl{</AddOnClick>}@*);
\end{lstlisting}

This elegantly hides channel actions, but at the same time,
we cannot make excessive assumptions about the 
event (\texttt{onClick}) and UI element (\texttt{<button>})
that the send action will be attached to --
these should be left as implementation details that 
the developer can customise for their application.

Instead, we present a novel approach for binding
channel actions to UI elements in a way that \textit{never
exposes channel resources}:

\begin{enumerate}
\item
The abstract send state React component defines
\textbf{factory properties}
that allow the developer to \textit{generate React components}
for performing send actions, such that the API lets the developer
specify which UI event should trigger the channel action,
and the payload to send. It defines one factory property per
permitted send action at that state.

\item
The runtime provides a \textit{factory method}
which allows the abstract send state component to
generate the aforementioned factory properties
for the channel actions supported at that state.
\end{enumerate}

We walk through the implementation details by working backwards:
showing how the developer would use the API first,
before deriving the type signatures that we need to
generate for the abstract send state component inherited by the developer.

\subsubsection{Developer API}

We show the developer's implementation 
of a send state for the \trole{Client} endpoint
from the \tprotocol{Adder} protocol in \cref{lst:adderclientsenddev}.
Component state is an implementation detail 
that is not related to EFSM execution, so we allow
the developer to customise this through specifying an 
interface (\cref{line:clientsendstatestate}).

\begin{figure}[!h]
\begin{lstlisting}[language=javascript,tabsize=2]
type State = { num1: number, num2: number, message: string }; (*@\label{line:clientsendstatestate}@*)
class InputWindow extends S40<State> {
	render() {
		const { num1, num2, message } = this.state;
		const (*@\hl{Add}@*) = this.ADD('onClick', (*@\label{line:clientsendfactory1}@*)
			(ev: React.MouseEvent) => [num1, num2]); (*@\label{line:clientsendpayload1}@*)
		const (*@\hl{Quit}@*) = this.QUIT('onMouseOver', (*@\label{line:clientsendfactory2}@*)
			(ev: React.MouseEvent) => [message]); (*@\label{line:clientsendpayload2}@*)
		return (<div>
			// Omitting <input>s for entering numbers, message
			(*@\hl{<Add>}@*)<button>Sum</button>(*@\hl{</Add>}@*)
			(*@\hl{<Quit>}@*)<button>Quit</button>(*@\hl{</Quit>}@*)
		</div>);
	}
}
\end{lstlisting}
\captionof{lstlisting}{Developer Implementation for \trole{Client}
Send State in \tprotocol{Adder} protocol}
\label{lst:adderclientsenddev}
\end{figure}

The factory API for binding send actions to UI elements appear on
\cref{line:clientsendfactory1,line:clientsendfactory2}.
For example, \cref{line:clientsendfactory1} reads: 
\textit{build a React component that sends the \tmsg{ADD}
message with \texttt{[num1, num2]} as payload,
when the user clicks on it}.
\tmsg{ADD} is the factory API we generate: it takes an event identifier
and an event handler which must return the payload 
-- \textit{appropriately typed with respect to the Scribble protocol} -- 
for sending the \tmsg{ADD} message.
Swapping \cref{line:clientsendpayload1,line:clientsendpayload2}
will result in a compile-time type error due to invalid 
payload types, which is the intended behaviour.

We will define the factory API implementation in a way that
calls the event handler to obtain the payload,
sends the payload message, and transitions to the successor state,
such that the send action is never performed more than once.

\subsubsection{Typing the Abstract Send Component}
Referring back to \cref{lst:adderclientsenddev},
a first attempt for typing \tmsg{ADD} could be

\begin{lstlisting}[language=javascript,numbers=none]
(ev: string, handler: (e: any) => [number, number]) 
	=> Constructor<React.Component>
\end{lstlisting}

However, the type of \texttt{ev} depends on the value of
\texttt{event} -- for example, an \texttt{onClick} event
is handled by a \texttt{React.MouseEvent} handler.
React defines the supported event handlers under the
\texttt{React.DOMAttributes} interface (\cref{lst:reactdomattrs}).
We can use an index type query to express the dependency,
and type inference in conditional types\footnote{
We can \lstonelinejs{infer} type variables inside the
\lstonelinejs{extends} clause of a conditional type.
Specifically, \lstonelinejs{FunctionArguments<T> = 
T extends (...args: infer R) ? R : never}.
}
to extract the function argument types:

\begin{lstlisting}[language=javascript,numbers=none]
<K extends keyof React.DOMAttributes>
(ev: K, handler: (e: FunctionArguments<React.DOMAttributes[K]>) 
		=> [number, number])
	=> Constructor<React.Component>)
\end{lstlisting}

Contrary to our discussions about modelling dependent types in
\cref{subsection:nodeefsmimplementation}, we can do this here
because: \textbf{(1)} we are using index type queries instead
of conditional types, and \textbf{(2)} we know the type
of \texttt{ev} at compile-time from the developer's implementation,
so we are not resolving index type queries with union types.

\begin{figure}[!h]
\begin{lstlisting}[language=javascript,numbers=none]
interface DOMAttributes<T> {
	children?: ReactNode; (*@\label{line:reactdomattrsnonfunc}@*)
	onChange?: FormEventHandler<T>;
	onClick?: MouseEventHandler<T>;
	...
};
\end{lstlisting}
\captionof{lstlisting}{Snippet of the \texttt{React.DOMAttributes} Interface}
\label{lst:reactdomattrs}
\end{figure}

However, \texttt{React.DOMAttributes} also contains
non-function properties -- on \cref{lst:reactdomattrs}
\cref{line:reactdomattrsnonfunc},  ``children'' is not a valid event.
We define utility types under Types.ts 
to extract function properties from an interface 
(\cref{lst:extractfuncprops}): we first extract
the \textit{names} of function properties into an union type 
by mapping property names of non-function properties to 
\lstonelinejs{never} (\cref{line:extractfuncprops1})
and performing an index type query (\cref{line:extractfuncprops2})
to obtain an union type of function property names\footnote{
The union type technically contains \lstonelinejs{never}
as the names for non-function properties, but 
TypeScript removes \lstonelinejs{never} from union types
in the same way that $A \vee \bot \equiv A$.
}
which removes all \lstonelinejs{never} constituents;
then extract the properties from the interface that are
indexed by the filtered names (\cref{line:extractfuncprops3})

\begin{figure}[!h]
\begin{lstlisting}[language=javascript,tabsize=2]
type FunctionPropertyNames<T> = {
	[K in keyof T]: T[K] extends Function | undefined ? K : never; (*@\label{line:extractfuncprops1}@*)
}[keyof T]; (*@\label{line:extractfuncprops2}@*)
type FunctionProperties<T> = Pick<T, FunctionPropertyNames<T>>; (*@\label{line:extractfuncprops3}@*)
type DOMEvents = FunctionProperties<React.DOMAttributes<any>>;
\end{lstlisting}
\captionof{lstlisting}{Extracting Function Properties from 
TypeScript Interface}
\label{lst:extractfuncprops}
\end{figure}

We construct a generic type parameterised on the payload type, 
which lets us define the \tmsg{ADD} and \tmsg{QUIT}
factory property easily.

\begin{lstlisting}[language=javascript,numbers=none]
type EventHandler<Payload, K extends keyof DOMEvents> = 
	(event: FunctionArguments<DOMEvents[K]>) => Payload;
	
type SendComponentFactory<Payload> = <K extends keyof DOMEvents>
	(event: K, handler: EventHandler<Payload, K>) => 
		Constructor<React.Component>;
\end{lstlisting}

The factories are instantiated by the abstract class
as \lstonelinejs{protected} properties, in order to allow 
access by the developer's subclass implementation.
Channel resources are managed by the runtime, so
we need the runtime to pass a ``higher-order factory''
(a factory method that generates a factory)
which lets the send state component generate the required
component factories (i.e. \tmsg{ADD} and \tmsg{QUIT})
with channel actions pre-injected into the component.
We discuss the runtime in \cref{section:reactruntime},
but for now, it is sufficient to understand that
the generated send component abstract class receives
the higher-order factory as a prop, and that this
prop binds the channel action into the factory that it returns.

Now we are in a position to accurately type the abstract
send component \texttt{S40} in \cref{lst:reactsendcomponent}.
We hint at the usage of the higher-order factory on
\cref{line:factory1,line:factory2}:
just using \cref{line:factory1} as an example,
the send component asks the runtime to generate a 
component factory for sending a message 
labelled \lstonelinejs{'ADD'}
with payload typed
\lstonelinejs{[number, number]},
then transition to the \texttt{ReceiveState.S42}
successor state.

\begin{figure}[!h]
\begin{lstlisting}[language=javascript,tabsize=2,title=S40.tsx]
type Props = { factory: SendComponentFactoryFactory };
abstract class S40<State> extends React.Component<Props, State> {
	protected ADD: SendComponentFactory<[number, number]>;
	protected QUIT: SendComponentFactory<[string]>;
	
	constructor(props: Props) {
		super(props);
		this.ADD = props.factory<[number, number]>( (*@\label{line:factory1}@*)
			'ADD', ReceiveState.S42);
		this.QUIT = props.factory<[string]>( (*@\label{line:factory2}@*)
			'QUIT', ReceiveState.S43);
	}
}
\end{lstlisting}
\captionof{lstlisting}{Generated Type for \trole{Client} Send State
in \tprotocol{Adder} protocol}
\label{lst:reactsendcomponent}
\end{figure}

\subsection{Receive States}

We also encode receive states as abstract React components,
and leverage props to ensure channel resources are not exposed
to the developer.

According to the EFSM, a receive state means
the endpoint is ``waiting for the message'',
and the actual receive event represents a transition
to a successor state.
This means that the developer's implementation will define
what to render on the DOM when the endpoint is waiting
for the message, rather than updating the DOM with the
received message.
Referring to \cref{fig:adderclientfsm} again,
S42 will be rendered when the endpoint is waiting for
\tmsg{RES}, but once received, S40 will be immediately
rendered. The developer will need to manage the flow
of stateful application data between UI components separately
-- \fancyname{ReactMPST} only generates code for managing
the communication automaton.

We generate an \textit{abstract method} 
for each permitted receive
transition, which is used as the receive handler.
This allows the developer to intercept the
message and perform operations -- such as updating
application state -- before the successor state is rendered.
Specifying receive handlers as abstract methods mandate
the developer to provide a concrete implementation.

We demonstrate below how the developer could implement
receive state S42 -- here, the developer manages
stateful application data using the
\textit{sessionStorage} API \cite{SessionStorage}
available on the browser.

\begin{lstlisting}[language=javascript]
class WaitScreen extends S42 {
	render() { return <h1>Waiting</h1>; }
	RES(n: number) {
		sessionStorage.setItem('result', n);
	}
}
\end{lstlisting}

Receive states in \fancyname{ReactMPST}
work in a similar way as with \fancyname{NodeMPST}.
The runtime will invoke the \tmsg{RES} method
implemented by the developer upon receiving the message
through the WebSocket.
In order to do so, the S42 component 
needs to \textit{register}
the handler with the runtime.

Recall that React supports an uni-directional data flow,
so there is no direct way to pass the handler
back up to the parent component.
We overcome this limitation by allowing the
runtime to pass a \texttt{register} function to the
receive state component as a prop; once the receive state
component is mounted on the DOM, it proactively 
registers the handlers with the runtime.

This raises questions about the type signature for the
\texttt{register} function, as the receive handler
identifiers and type signatures differ between receive states.
We adopt the same strategy as with \fancyname{NodeMPST}:
the receive handler registered by the receive state
to the runtime also performs the message parsing,
so all receive states register a ``general handle function'' typed

\begin{lstlisting}[language=javascript,numbers=none]
type MessageHandler = (message: any) => void;
\end{lstlisting}

and define their own \texttt{MessageHandler} to handle
state-specific message parsing and processing.
We construct similar types for the state-specific
labels and message structures.
This is illustrated in \cref{lst:reactreceivetypes}) --
we use state S43 from \cref{fig:adderclientfsm} hereafter
to show how our approach scales with multiple branches.

\begin{figure}[!h]
\begin{lstlisting}[language=javascript,tabsize=2]
enum Labels { THANKS = 'THANKS', TERMINATE = 'TERMINATE' };
interface THANKSMessage {
    label: Labels.THANKS,
    payload: [string],
};
interface TERMINATEMessage {
    label: Labels.TERMINATE,
    payload: [],
};
type Message = | THANKSMessage | TERMINATEMessage;
\end{lstlisting}
\captionof{lstlisting}{Generated Label and Message Types
in \fancyname{ReactMPST}}
\label{lst:reactreceivetypes}
\end{figure}
	
We generate the \texttt{handle()} method for each
abstract receive state component for state-specific message
parsing and processing. We show the full generated code
for state S42 in \cref{lst:reactreceivecomponent}.
Unlike \fancyname{NodeMPST},
the \texttt{handle()} method (\cref{line:reactreceivehandle})
returns the successor state identifier to the runtime
to make the transition.
We leverage the discriminated union from our
generated message types (\cref{lst:reactreceivetypes})
to invoke the correct receive handler (\cref{line:reactreceivediscriminate}).

Again, due to language limitations, we are unable to 
simplify the message processing step as

\begin{lstlisting}[language=javascript,numbers=none]
this[parsedMessage.label](...parsedMessage.payload);
\end{lstlisting}

because this is implicitly expressing a type dependency
over union types,
and \texttt{parsedMessage.label} is unknown at compile-time.

\begin{figure}[!h]
\begin{lstlisting}[language=javascript,tabsize=2]
type Props = {
	register: (handle: ReceiveHandler) => void
};
abstract class S42<S> extends React.Component<Props, S> {
	abstract THANKS(payload1: string): void;
	abstract TERMINATE(): void;
	
	componentDidMount() {
		this.props.register(this.handle.bind(this));
	}
	
	handle(message: any): State { (*@\label{line:reactreceivehandle}@*)
		const parsedMessage = JSON.parse(message) as Message;
		switch (parsedMessage.label) { (*@\label{line:reactreceivediscriminate}@*)
			case Labels.THANKS:
				this.THANKS(...parsedMessage.payload);
				return TerminalState.S41;
			case Labels.TERMINATE:
				this.TERMINATE(...parsedMessage.payload);
				return TerminalState.S41;
		}
	}
}
\end{lstlisting}
\captionof{lstlisting}{Generated Code for \trole{Client} 
Receive State S42 in \tprotocol{Adder} protocol}
\label{lst:reactreceivecomponent}
\end{figure}

\subsection{Terminal States}
To be consistent with the conventions defined for \fancyname{NodeMPST},
the browser endpoint proactively closes the connection at its terminal state.
We generate an abstract React component for terminal states 
to allow the developer to customise what to
\texttt{render} upon protocol termination.

\begin{lstlisting}[language=javascript]
type Props = { terminate: () => void };
abstract class S41<State> extends React.Component<Props, State> {
	componentDidMount() { this.props.terminate(); }	
}
\end{lstlisting}

When the component is mounted on the DOM, it notifies the runtime
(via the received \texttt{terminate} prop) to terminate the session.