\section{Challenges}
\label{section:routerchallenges}

We need to formalise our concept of routing by extending session
type theory, such that, given a communication protocol implementing 
an arbitrary topology, we can encode it into our extended theory and
prove that safety properties (e.g. well-formedness, deadlock freedom)
are preserved, and that communication traces are preserved.
This must be done \textit{very carefully}, as
naive definitions of routing risk 
over-serialising the original communication,
thereby not preserving all communication traces.

\begin{example}[Naive definition of routing in global types]
Consider the global type
\[
A \to B: \tmsg{M1}. S \to B: \tmsg{M2}. \texttt{end}
\]
By existing LTS semantics over
global types introduced in 
Definition 3.2 of \cite{characterisation}, 
then $\aout{S}{B}{\tmsg{M2}}$
is a prefix of a valid execution trace.
If we define our routing construct simply by replacing every single
$A \to B: \tmsg{M}$ interaction with $A \to S: \tmsg{M}.
S \to B: \tmsg{M}$, we transform the global type into
\[
A \to S: \tmsg{M1}. S \to B: \tmsg{M1}. S \to B: \tmsg{M2}. \texttt{end}
\]
our naive definition of routing has over-serialised the protocol
since $\aout{S}{B}{\tmsg{M2}}$ can no longer occur first because \trole{S}
is forced to send \tmsg{M1} first, which it cannot do so until receiving
from \trole{A}.
This implies that we need to handle routing interactions differently
from ``normal'' send/receive actions.
\end{example}

There is existing work \cite{LinearDecomp, BinaryDecomp, BinaryDuality}
that resembles our routing proposal through
\textit{decomposing} multiparty sessions into a collection
of binary sessions,
but these approaches
specify an endpoint to exclusively carry out
routing responsibilities and act as a centralised orchestrator 
for the multiparty session.
Our proposal recognises that, in the context of web applications,
\textit{there is a need for server-side endpoints to 
implement business logic},
so enforcing the server to be an exclusive orchestrator makes our work
incompatible with web service protocols; 
we also argue that the server can still participate in the 
multiparty session and concurrently perform its routing duties
in a way that is transparent to non-server endpoints, and crucially,
does not overserialise the permitted interactions. 
We introduce our extension as \newtheory,
and proceed to prove these claims in \cref{chap:theory}.

Likewise, the implementation of routed session types 
in \codegen needs to accurately reflect the theory.
It also needs to be 
as transparent to the developer as possible to minimise
the learning curve overhead demanded 
by complicating the generated APIs.
We document our extensions
to \codegen in \cref{chap:impl}.