\chapter{Lemmas and Proofs}

\section{Lemmas and Proofs for Chapter 8}
\label{section:proofs}

\begin{lemma}[Local LTS Preserves Merge]
Let $T_1, T_2$ be local types.
Suppose $\tmerge{T_1}{T_2}$ exists.

\[
(\treduce{T_1}{T'_1}{l}) \wedge (\treduce{T_2}{T'_2}{l})
	\Longrightarrow
(\tmerge{T'_1}{T'_2}) \text{ exists}
\]

\label{lem:localltspreservemerge}
\end{lemma}

\begin{proof}
By simultaneous induction on
$\tmerge{T_1}{T_2}$, $\treduce{T_1}{T'_1}{l}$,
and $\treduce{T_2}{T'_2}{l}$.

\end{proof}

\begin{lemma}[Projection and Participation]

\[
\proj{G}{p} = \tend \Longleftrightarrow \mrole{p} \notin \pt{G}
\]

\label{lem:projpt}
\end{lemma}

\begin{proof} 
Prove $(\Longrightarrow)$ by induction on the structure of $G$.
Prove $(\Longleftarrow)$ using the contrapositive (stated below)
by induction on the derivation of $\pt{G}$.

\[
\mrole{p} \in \pt{G} \Longrightarrow \proj{G}{p} \neq \tend
\]
\end{proof}

\begin{lemma}[Commutativity between Encoding and Substitution]
Let $G, G'$ be global types, and $\mrole{s}$ be a role.

\[
\enc{G[G' / \trecvar]}{s} = \enc{G}{s}\left[ \enc{G'}{s} / \trecvar \right]
\]

\label{lem:encsub}
\end{lemma}

\begin{proof}
By induction on the structure of $G$.
\end{proof}

\begin{lemma}[???]
\[
\forall \mrole{s}. ~ \pt{G} \subseteq \pt{\enc{G}{s}}
\]
\label{lem:encpreservept}
\end{lemma}

\begin{proof}
The following is logically equivalent.

\[
\forall \mrole{r}, \mrole{s}. ~ \left(
\mrole{r} \in \pt{G} \Longrightarrow \mrole{r} \in \pt{\enc{G}{s}}
\right)
\]

We prove this by induction on the structure of $G$.
\end{proof}

\begin{lemma}[Encoding Preserves Privacy]
The encoding on global types will not introduce
non-server roles that were not participants of the 
original communication.

\[
\forall \mrole{r}, \mrole{s}, G. ~ \left(
\mrole{r} \neq \mrole{s} \wedge \mrole{r} \notin \pt{G}
\Longrightarrow \mrole{r} \notin \pt{\enc{G}{s}}
\right)
\]
\label{lem:encprivacy}
\end{lemma}

\begin{proof}
The following is logically equivalent.
\[
\forall \mrole{r}, \mrole{s}, G. ~ \left(
\mrole{r} \neq \mrole{s} \wedge \mrole{r} \notin \pt{G}
\Longrightarrow \mrole{r} \notin \pt{\enc{G}{s}}
\right)
\]

We prove this by induction on the structure of $G$,
assuming that $\mrole{r} \neq \mrole{s}$ for arbitrary
roles $\mrole{r}$, $\mrole{s}$.
\end{proof}

\begin{lemma}[Correspondence between Encodings on Global Types
and Local Types]

\[
\forall \mrole{r}, \mrole{s}, G. ~ \left(
\mrole{r} \neq \mrole{s} 
	\Longrightarrow
\proj{\enc{G}{s}}{r} = \enclocal{\proj{G}{r}}{r}{s}
\right)
\]

\label{lem:enclink}
\end{lemma}

\begin{proof}
By induction on the structure of $G$,
\cref{lem:encpreservept,lem:encprivacy}.
\end{proof}

\begin{lemma}[Local Type Encoding Preserves Equality of Projection]
\[
\forall G_1, G_2, \mrole{r}, \mrole{s}. ~ \left(
(\proj{G_1}{r}) = (\proj{G_2}{r}) \wedge \mrole{r} \neq \mrole{s}
	\Longrightarrow
\proj{\enc{G_1}{s}}{r} = \proj{\enc{G_2}{s}}{r}
\right)
\]

\label{lem:encprojeq}
\end{lemma}

\begin{proof} By consequence from \cref{lem:enclink}.

Take $G_1, G_2, \mrole{r}, \mrole{s}$ arbitrarily. 
Assume $(\proj{G_1}{r}) = (\proj{G_2}{r})$ and $\mrole{r} \neq \mrole{s}$.

We need to show $\proj{\enc{G_1}{s}}{r} = \proj{\enc{G_2}{s}}{r}$,
but by \cref{lem:enclink}, it is sufficient to show
\[
\enclocal{\proj{G_1}{r}}{r}{s} = \enclocal{\proj{G_2}{r}}{r}{s}
\]

Define an ``inner'' function $f(z) = \enclocal{z}{r}{s}$.

By definition, $\forall x, y. (x = y \Longrightarrow f(x) = f(y))$.

We have $(\proj{G_1}{r}) = (\proj{G_2}{r})$ by assumption,
so we can conclude.

\begin{align*}
(\proj{G_1}{r}) = (\proj{G_2}{r}) 
&\Longrightarrow f(\proj{G_1}{r}) = f(\proj{G_2}{r})  \\
&\Longrightarrow
\enclocal{\proj{G_1}{r}}{r}{s} = \enclocal{\proj{G_2}{r}}{r}{s}
\end{align*}

\end{proof}

\begin{lemma}[Encoding on Global Types Preserves Merge]
Take global types $G_1, G_2$
and roles $\mrole{r}, \mrole{s}$ such that $\mrole{r} \neq \mrole{s}$.
Suppose $\proj{G_1}{r}$ and $\proj{G_2}{r}$ exist.

\[
\tmerge{(\proj{G_1}{r})}{(\proj{G_2}{r})} \text{ exists}
	\Longrightarrow
\tmerge{(\proj{\enc{G_1}{s}}{r})}{(\proj{\enc{G_2}{s}}{r})} 
	\text{ exists}
\]

\label{lem:encglobalpreservemerge}
\end{lemma}

\begin{proof}
By induction on the structure of $\tmerge{T_1}{T_2}$,
\cref{lem:projpt,lem:encprojeq}

\end{proof}