\begin{abstract}
Advancements in mobile device compute power have made interactive web
applications possible, allowing the web browser to render contents dynamically
and support low-latency communication with the server.
This comes at a cost to the developer, who now needs to reason more about the
communication correctness of their application as they support more complex
communication patterns.

Multiparty session types (MPST) provide a framework for verifying conformance
of implementations to their prescribed communication protocol.
Existing proposals for applying the MPST framework in application development
either neglect the event-driven nature of web applications, or lack
compatibility with industry tools and practices, which discourages mainstream
adoption by web developers.

In this paper, we present an implementation of the MPST framework for
developing interactive web applications using familiar industry tools in
TypeScript and the \textit{React.js} framework.
The developer can use the Scribble protocol
language to specify the protocol and use the Scribble toolchain to validate
and obtain the local protocol for each role.
The local protocol describes the interactions of the global communication
protocols observed by the role.
We encode the local protocol into TypeScript types, catering for server-side
and client-side targets separately, and show that our encoding guarantees that
only implementations which conform to the protocol can type-check.
We demonstrate the effectiveness of our approach through a web-based
implementation of the classic \textit{Noughts and Crosses} game from a MPST
formalism of the game logic.

\end{abstract}

